---
title: 'MC Integration'
author: "Qinqing Li, s2024889"
output:
  pdf_document:
    number_sections: yes
  html_document:
    number_sections: yes
header-includes:
  - \newcommand{\bm}[1]{\boldsymbol{#1}}
  - \newcommand{\mat}[1]{\begin{bmatrix}#1\end{bmatrix}}
---

```{r setup, include = FALSE}
# Modify this setup code chunk to set options
# or add extra packages etc if needed.
# See the project instructions for more details
# on what code to show, and where/how.
library(ggplot2)
library(dplyr)
# Set default code chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE
)
```

# Overdispersed relative to Poisson distribution

approach 1
```{r}
# Vectorised over m
mu <- log(2) - 1/2
K <- 10000
m <- 0:15
P_Y <- numeric(length(m))
for (loop in seq_len(K)) {
  x <- rnorm(1, sd = 1)
  P_Y <- P_Y + dpois(m, lambda = exp(mu + x))
}
P_Y <- P_Y / K
print(P_Y)


suppressPackageStartupMessages(library(ggplot2))
ggplot(data.frame(m = m,
                  P_Y = P_Y,
                  P_Poisson = dpois(m, lambda = 2))) +
  geom_point(aes(m, P_Y, col = "MC")) +
  geom_point(aes(m, P_Poisson, col = "Poisson")) +
  geom_line(aes(m, P_Y, col = "MC")) +
  geom_line(aes(m, P_Poisson, col = "Poisson"))
```

approach 2
```{r}
K <- 10000
m <- 0:15
mu <- log(2) - 1/2
for (i in seq_along(m)) {
  x <- rnorm(K, sd = 1)
  total <- sum(dpois(m[i], lambda = exp(mu + x)))/K
  print(total)
}

suppressPackageStartupMessages(library(ggplot2))
ggplot(data.frame(m = m,
                  P_Y = P_Y,
                  P_Poisson = dpois(m, lambda = 2))) +
  geom_point(aes(m, P_Y, col = "MC")) +
  geom_point(aes(m, P_Poisson, col = "Poisson")) +
  geom_line(aes(m, P_Y, col = "MC")) +
  geom_line(aes(m, P_Poisson, col = "Poisson"))
```

```{r}
doverpois <- function(m, mu, sigma, K) {
  total <- numeric(length(m))
  
  for (i in seq_along(m)) {
    x <- rnorm(K, sd = 1) 
    total[i] <- sum(dpois(m[i], lambda = exp(mu + x))) / K
  }
  data.frame(
    m = m,
    P_Y = total,
    P_Poisson = dpois(m, lambda = exp(mu + sigma^2 / 2))
  )
}

```

```{r}
ggplot(doverpois(m = 0:30, mu = log(8) - 0.125, sigma = 0.5, K = 10000))
suppressPackageStartupMessages(library(ggplot2))
ggplot(doverpois(m = 0:30, mu = log(8)-0.125, sigma = 0.5, K = 10000)) +
  geom_point(aes(m, P_Y, col = "MC")) +
  geom_point(aes(m, P_Poisson, col = "Poisson")) +
  geom_line(aes(m, P_Y, col = "MC")) +
  geom_line(aes(m, P_Poisson, col = "Poisson"))
```

# Archaeology in the Baltic sea - using Monte Carlo integration

The aim of this simulation is to estimate the true value of N, which represents the total number of people buried, and ϕ, the probability of finding a femur (left or right).

Our prior belief is that there are approximately $1000$ individuals buried, and the probability of finding a femur (either left or right) is $0.5$. The prior distribution is $N~Geom(\xi)$ where $\xi = \frac{1}[1+1000}$, and $\phi ~ beta(2,2)$.

```{r}
arch_loglike <- function(y, dataframe) {
  y1 <- y[1]
  y2 <- y[2]
  N <- dataframe$N
  phi <- dataframe$phi
  loglikelihood <- numeric(length(N))  

  for (i in seq_along(N)) {
    loglike <- -lgamma(y1 + 1) - lgamma(y2 + 1) -
               lgamma(N[i] - y1 + 1) - lgamma(N[i] - y2 + 1) + 
               2 * lgamma(N[i] + 1) +
               (y1 + y2) * log(phi[i]) + 
               (2 * N[i] - y1 - y2) * log(1 - phi[i])
    loglikelihood[i] <- loglike
  }

  return(loglikelihood)
}
```



```{r}
estimate <- function(y, xi, a, b, K) {
  set.seed(123)
  samples <- data.frame(
    N = rgeom(K, prob = xi),
    phi = rbeta(K, shape1 = a, shape2 = b)
  )
  loglike <- arch_loglike(y = y, dataframe = samples)
  p_y <- mean(exp(loglike))
  c(p_y = p_y,
    E_N = mean(samples$N * exp(loglike) / p_y),
    E_phi = mean(samples$phi * exp(loglike) / p_y))
}
```

```{r}
estimate(y = c(237, 256), xi = 1/1001, a = 0.5, b = 0.5, K = 10000)
```

We conclude that there are approximately 982 individuals buried, and the probability of uncovering a femur is 0.39.